The Second Data Release of the INT Photometric H$\alpha$ Survey of the Northern Galactic Plane (IPHAS DR2)

### 2014-04-10 paper/iphas-dr2.tex
-

\begin{figure*}
\captionsetup[subfigure]{labelformat=empty}
\begin{subfigure}[b]{0.45\linewidth}
\centering
\includegraphics[width=\textwidth]{./plots/pairmag/pairmag.pdf}
\caption{}
\label{fig:pairmag}
\end{subfigure}
\hspace{0.5cm}
\begin{subfigure}[b]{0.45\linewidth}
\centering
\includegraphics[width=\textwidth]{./plots/pairmag/pairmag-reliable.pdf}
\caption{}
\label{fig:pairmag_reliable}
\end{subfigure}
\caption{Photometric repeatability as a function of magnitude
         for both all sources in the catalogue (left panel)
         and for the \emph{veryReliable} sources alone (right panel).
         The values shown are the mean residuals
         between the primary and the secondary detections,
         while the error-bars show the standard deviations.
         Left panel: 
         the best photometric repeatability is reached at $r=14$
         with $\sigma_{r-r2}~=~0.041$~mag.
         Bright stars at $r<13$ and $i<12$ 
         show increasing uncertainties due to saturation effects.
         Right panel: we find that applying the quality criteria
         has improved the photometric repeatability significantly.
         The best repeatability is again reached at $r=14$
         but has reduced to $\sigma=0.028$ mag.
         The quality criteria have also been successful
         at removing objects with large systematics at the bright
         and faint ends.
    }
\end{figure*}

@@

\subsection{Source densities and blending}
\label{sec:densities}

The mean source density as a function of Galactic longitude
is shown in Fig.~\ref{fig:density} (thick blue line).
The densities were computed by counting the 
number of sources in $1^\circ$-wide longitude bins,
each covering the entire latitude range $-5^\circ<b<+5^\circ$.
Unsurprisingly, we find the average source densities to increase
towards the Galactic centre.
For example, the average source density near $l\simeq 30^\circ$
is roughly 30\,000 objects per square degree,
which is five times more than the density
found near $l\simeq 180^\circ$.

In addition to the global trend, 
there are significant variations
in the source density on smaller scales.
For example,  we find a significant drop near the constellations 
of Aquila ($l\simeq40$) and Cygnus ($l\simeq80$ and $l\simeq90$),
which are regions known to be affected
by high levels of foreground extinction.
In fact dark clouds are readily visible in the night sky
towards these constellations,
and they are often referred to as the `the Great Rift' \citep[e.g.][]{BokBok}.

We warn that the relative densities reported here 
have not been corrected for survey completeness
or differences in the observing conditions across the survey fields.
For example, the dip in the density near $l\simeq140^\circ$
is an artificial feature caused by gaps
in the footprint coverage (seen in Fig.~\ref{fig:footprint}).
In a forthcoming paper,
we will calibrate the IPHAS-based source densities
by injecting artificial stars into the IPHAS images
and measuring their recovery rate (Farnhill et al., in preparation).
IPHAS has the potential to offer
calibrated, two-dimensional stellar density maps
which can be used to constrain detailed models of our Galaxy.

\begin{figure*}
    \includegraphics[width=\textwidth]{./plots/density/density.pdf} 
    \caption{Mean number density of sources in the catalogue 
    as a function of Galactic longitude, 
    with and without blended sources included. 
    The densities shown were computed by counting the sources 
    at each longitude between $-5^\circ<b<+5^\circ$ (upper blue line).
    We also show the densities based on only counting those sources 
    for which the \emph{deblend} flag is {\sc false}, 
    i.e. unconfused sources for which the CASU pipeline did not have to apply 
    a deblending procedure (lower red line). 
    }
    \label{fig:density}
\end{figure*}

In Fig.~\ref{fig:density} we also shows the density
of non-blended sources (thin red line).
These are sources for which the \emph{deblend} flag is {\sc false},
i.e. sources for which the CASU pipeline did not have to apply 
a deblending procedure to separate the flux
originating from two or more overlapping objects.
We find that the fraction of sources affected by blending problems
is strongly correlated with the local source density.
For example, only 11\% of the sources are blended
at $l>90^\circ$, whereas 24\% are blended at $l<90^\circ$.

As we explained earlier, blended sources must be used with caution.
In particular, we warn that blended objects
are likely candidates to have fallen victim
to source confusion during the band-merging procedure.
We have started investigating the use of
more advanced PSF-fitting routines
in which sources are measured simultaneously across all bands,
perhaps guided by an external astrometric catalogue such as the one
that is to be provided by Gaia.

@@

A new data release for the INT/WFC Photometric \ha\ Survey
of the Northern Galactic Plane (IPHAS) was presented.
It is the first survey to offer comprehensive CCD photometry
of point sources across the Galactic Plane at visible wavelengths,
taking in the entire Northern Milky Way at $|b|<5^\circ$.
The new 97-column catalogue provides single-epoch photometry
across 92\% of the survey area,
and is the first quality-controlled and uniformly calibrated
catalogue to have been constructed from the imaging data.
We explained the data reduction and quality control procedures that
were applied,
described and tested the global photometric calibration,
and detailed the construction of the source catalogue.

The observations included in this release
achieved a median seeing of 1.1 arcsec
and a $5\sigma-$depth of $r=21.2\pm 0.5$, $i=20.0\pm 0.3$, \ha$=20.3\pm 0.3$.
The global calibration and photometric repeatability
were found to be accurate at the level of $0.03$ mag (rms),
providing a significant improvement over the 
previous data release.
The source catalogue provides the best-available
single-epoch astrometry and photometry
for 219~million unique sources.
We detailed a list of recommended quality criteria
to select objects with reliable colours from the catalogue.
Finally, we demonstrated the use of the survey's unique
$(r-\ha,\ r-i)$ diagram for characterising stellar populations
and selecting emission-line objects.

In future work, we intend to draw upon the PanSTARRS photometric
survey to further improve the accuracy of our global calibration.
We also aim to correct the IPHAS photometry for the
radial field distortions,
plan to produce a catalogue 
which details all the secondary detections
to aid time-domain studies,
and are investigating the use of photometry
derived from PSF-fitting across the survey.

>>>>

\begin{figure*}
\captionsetup[subfigure]{labelformat=empty}
\begin{subfigure}[b]{0.45\linewidth}
\centering
\includegraphics[width=\textwidth]{figures/repeatability/pairmag.pdf}
\caption{}
\label{fig:pairmag}
\end{subfigure}
\hspace{0.5cm}
\begin{subfigure}[b]{0.45\linewidth}
\centering
\includegraphics[width=\textwidth]{figures/repeatability/pairmag-reliable.pdf}
\caption{}
\label{fig:pairmag_reliable}
\end{subfigure}
\caption{Photometric repeatability as a function of magnitude
         for both all sources in the catalogue (left panel)
         and for the \emph{veryReliable} sources alone (right panel).
         The values shown are the mean absolute residuals
         between the primary and the secondary detections,
         while the error-bars show the standard deviations.
         Left panel: 
         the best photometric repeatability is reached at $r=14$
         with $\sigma_{r-r2}~=~0.041$~mag.
         Bright stars at $r<13$ and $i<12$ 
         show increasing uncertainties due to saturation effects.
         Right panel: we find that applying the quality criteria
         has improved the photometric repeatability significantly.
         The best repeatability is again reached at $r=14$
         but has reduced to $\sigma=0.028$ mag.
         The quality criteria have also been successful
         at removing objects with large systematics at the bright
         and faint ends.
    }
\end{figure*}

@@

\subsection{Source counts and blending}
\label{sec:densities}

The number of sources in our catalogue,
counted in 1-by-10~deg$^2$ strips,
is shown in Fig.~\ref{fig:sourcecount}
as a function of Galactic longitude (thick blue line).	
Unsurprisingly, we find the number of sources
to increase towards the Galactic centre.
For example, the average source density near $l\simeq 30^\circ$
is roughly 300\,000 objects per square degree,
which is five times more than the density
found near $l\simeq 180^\circ$.

In addition to the global trend,
variations are also apparent on smaller scales.
For example, we find a significant drop near the constellations 
of Aquila ($l\simeq40$) and Cygnus ($l\simeq80$ and $l\simeq90$),
which are regions known to be affected
by high levels of foreground extinction
\citep[known as `the Great Rift', e.g.][]{BokBok}.

We warn that the source counts reported here 
have not been corrected for survey completeness
or variations in the depth of different survey fields.
For example, the dip in the density near $l\simeq140^\circ$
is an artificial feature caused by gaps
in the footprint coverage (seen in Fig.~\ref{fig:footprint}).
In a forthcoming paper,
we will calibrate the IPHAS-based source densities
by injecting artificial stars into the images
and measuring their recovery rates
(Farnhill et al., in preparation).
Such calibrated density maps may be used 
to constrain detailed models of our Galaxy.

\begin{figure*}
    \includegraphics[width=\textwidth]{figures/sourcecount/sourcecount.pdf} 
    \caption{Number of entries in the IPHAS DR2 source catalogue
    as a function of Galactic longitude.
    The upper blue line shows the number of sources
    counted in 1\deg-wide longitude bins.
    The lower red line uses the same binning
    but includes only those sources 
    for which the \emph{deblend} flag is {\sc false}, 
    i.e. unconfused sources for which the CASU pipeline
    did not have to apply a deblending procedure.
    In both cases only counted sources
    in the latitude range $-5^\circ<b<+5^\circ$,
    such that one may obtain approximate source densities
    by dividing the counts by 10~deg$^2$.
    The apparent variations in the source counts
    traces the structure of the Galaxy
    and the distribution of foreground extinction,
    but also includes instrumental effects
	such as variations in the survey depth
	and completeness (see text).
   }
    \label{fig:sourcecount}
\end{figure*}

In Fig.~\ref{fig:sourcecount} we also shows the number
of non-blended sources (thin red line).
These are sources for which the \emph{deblend} flag is {\sc false},
i.e. sources for which the CASU pipeline did not have to apply 
a deblending procedure to separate the flux
originating from two or more overlapping objects.
We find that the fraction of sources affected by blending problems
is strongly correlated with the local source density.
For example, only 11\% of the sources are blended
at $l>90^\circ$, whereas 24\% are blended at $l<90^\circ$.

As we explained earlier, blended sources must be used with caution.
In particular, we warn that blended objects
are likely to have fallen victim
to source confusion during the band-merging procedure.
We have started investigating the use of
more advanced PSF-fitting routines
in which sources are measured simultaneously across all bands,
which may be guided by an external astrometric catalogue
such as the one that is to be provided by Gaia.

@@

A new catalogue of 219 million sources has been
derived from the INT/WFC Photometric \ha\ Survey
of the Northern Galactic Plane.
It is the first catalogue to offer comprehensive CCD photometry
of point sources across the Galactic Plane at visible wavelengths,
taking in the Northern Milky Way at $|b|<5^\circ$.
The new 97-column catalogue provides single-epoch photometry
across 92\% of the survey area,
and is the first quality-controlled and uniformly calibrated
catalogue to have been constructed from the imaging data.
We explained the data reduction and quality control procedures that
were applied,
described and tested the global photometric calibration,
and detailed the construction of the source catalogue.

The observations included in this release
show a median seeing of 1.1 arcsec
and $5\sigma-$depths of $r=21.2\pm 0.5$, $i=20.0\pm 0.3$,
and \ha$=20.3\pm 0.3$.
The global calibration and photometric repeatability
were found to be accurate at the level of $0.03$ mag (rms),
providing a significant improvement over the 
previous data release.
The source catalogue provides the best-available
single-epoch astrometry and photometry
for 219~million unique sources.
We detailed a list of recommended quality criteria
to select objects with reliable colours from the catalogue.
Finally, we demonstrated the use of the survey's unique
$(r-\ha,\ r-i)$ diagram for characterising stellar populations
and selecting emission-line objects.

In future work, we intend to draw upon the PanSTARRS photometric
survey to further improve the accuracy of our global calibration.
We also plan to produce a catalogue 
which details all the secondary detections
to aid time-domain studies,
and are investigating the use of photometry
derived from PSF-fitting across the survey.
